The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2020-07-15 Paper/jaccardbaetica.tex
add new changes xavi review

The landscape of the \textit{Baetican} province was one of the best regions to face the increase in demand of olive oil across the Roman provinces. For this reason the area saw an increase in productivity as a massive infrastructure of olive oil production was gradually deployed. The production and distribution of olive oil increased from the first to the third centuries AD ~\citep{remesal_concierto}. Hundreds of presses and amphora-making workshops were build near large extensions covered by olive trees. The workshops where amphorae were made and filled with olive oil were places along the rivers Guadalquivir and Genil riverbanks (see Fig.\ref{workshop}).

@@

A large majority of this Baetican amphorae production is composed by Dressel 20 types\citep{dressel_ricerche_1878,millet_anforas_1998}. Dressel 20 amphorae have a recognisable globular form with short neck and oval-shaped handles. Its  capacity allowed to hold 70-75 litres of olive oil to be shipped and distributed throughout Empire \citep{berni_dressel_2016}.

@@

%\xavi{este título es muy raro, porque 1. que son casos de estudio ya lo sabes y 2. case study singular, consumption centres plural}

@@

Some authors argue elsewhere that this same \textit{Annona} could also have organized the supply of additional goods such as olive oil to the Roman legions \citep{remesal_annona_1986,remesal_annona_1990}\xavi{pero no hay evidencia? Las fuentes escritas no lo mencionan?}.

The importance of this olive oil trade is revealed by the massive amount of Dressel 20 amphorae found across the Roman provinces. This amphora has been commonly associated with the transportation of Baetican olive oil for supplying military camps and civil settlements \citep{berni_millet_epigrafianforica_2008}.\xavi{me parece repetitivo, porque unos párrafos antes ya hablabas de lo mismo. Quizás unificar texto aquí o allá?}

The military consumption of olive oil in Roman provinces may be related to two aspects: a) cultural consumption whose product is consumed by cultural reasons such as to identity or habit and b) economical reasons where olive oil distribution is based on the transport costs, unlike other products \citep[69-70]{carreras_britannia_1998}\xavi{no pillo esta frase}..

Olive oil supply seems to be particularly intense in militarised provinces that functioned as the borders of the Empire such as \textit{Britannia} and \textit{Germania}. The high percentage of Dressel 20 amphorae found in sites nearby military garrisons suggests the existence of diverse supply dynamics as there were trade processes flowing good from military sites to civil settlements\citep{remesal_annona_1986, carreras_britannia_1998}\xavi{lo que quieres decir es que tienes muchas ánforas en sitios civiles cerca de militares? no queda claro}.

@@

The evidence for olive oil consumption in \textit{Britannia} is scarce before the Roman conquest \citep{funari_corpus_1996,carreras_abastecimiento_2003}. Local population did not consume this product and it was not produced on the region as the landscape and climate of the British isles was not suitable for olive oil trees\citep[161]{monfort_britanniaen_1998}. Based on the archaeological record this absence of olive oil changed with the arrival of a large amount of Dressel 20 amphorae to \textit{Britannia} \citep[1]{carreras_britannia_1998}\xavi{en qué siglo?}. The demand of olive oil after the Roman conquest required of a new trade route due to the lack of local olive oil production; \textit{Baetica} was chosen as its main supply.

At this moment, we detect an increase of the olive oil exportation concurring with the displacement of legions during the military campaigns \citep[161]{monfort_britanniaen_1998}\xavi{repetitivo? fusionar párrafos}. This fact will have a particular intensity within sites close to the Hadrian Wall's garrisons. Olive oil production in \textit{Baetica} would cross the Atlantic until they reached the province and redistribute throughout the area from a series of strategic locations~\citep{carreras_atlantic_2012}. The increase of the exportation of Dressel 20 amphorae created an important commercial network for exchanges. Thus, the network was mainly focused on the support of soldiers during military campaigns\xavi{estas 2 frases repiten cosas ya explicadas}.

@@

According to previous studies those workshops became the most important centres of amphora production in the region\xavi{cita de estos estudios}. It is worth mentioning that a majority of these stamps were collected during ield surveys with no excavation, so this difference in sample size could be biased due to differences in intensity across the different sites~\citep{arva_1997}.

@@

The analysis of stamps in consumption areas also used the CEIPAC database and filtered in the same way than the previous dataset. We selected the centres with more or equal than five stamps\xavi{pero esto no se hizo en el de Baetica no? Por qué?}.

>>>>

The landscape of the \textit{Baetican} province was one of the best regions to face the increase in demand of olive oil across the Roman provinces. For this reason the area saw an increase in productivity as a massive infrastructure of olive oil production was gradually deployed. The production and distribution of olive oil increased from the first to the third centuries AD ~\citep{remesal_concierto}.

Hundreds of presses and amphora-making workshops were build near large extensions covered by olive trees. The workshops where amphorae were made and filled with olive oil were places along the rivers Guadalquivir and Genil riverbanks (see Fig.\ref{workshop}).

@@

The importance of this olive oil trade is revealed by the massive amount of Dressel 20 type amphorae found across the Roman provinces \citep{dressel_ricerche_1878,
millet_anforas_1998}. This amphora has been commonly associated with the transportation of Baetican olive oil for supplying military camps and civil settlements \citep{berni_millet_epigrafianforica_2008}.%\xavi{me parece repetitivo, porque unos párrafos antes ya hablabas de lo mismo. Quizás unificar texto aquí o allá?}.

Dressel 20 amphorae have a recognisable globular form with short neck and oval-shaped handles. Its capacity allowed to hold 70-75 litres of olive oil to be shipped and distributed throughout Empire \citep{berni_dressel_2016}.

%A large majority of this Baetican amphorae production is composed by Dressel 20 types\citep{dressel_ricerche_1878,
%millet_anforas_1998}.

@@

@@

Some hypotheses argue elsewhere that this same \textit{Annona} could also have organized the supply of additional goods such as olive oil to the Roman legions \citep{remesal_annona_1986,remesal_annona_1990}

%\xavi{pero no hay evidencia? Las fuentes escritas no lo mencionan? - He cambiado authors por hipotesis porque me quería referir a investigadores zorry}.

%The importance of this olive oil trade is revealed by the massive amount of Dressel 20 amphorae found across the Roman provinces. This amphora has been commonly associated with the transportation of Baetican olive oil for supplying military camps and civil settlements \citep{berni_millet_epigrafianforica_2008}.\xavi{me parece repetitivo, porque unos párrafos antes ya hablabas de lo mismo. Quizás unificar texto aquí o allá?} LO HE PUESTO MÁS ARRIBA!

Olive oil supply seems to be particularly intense in militarised provinces that functioned as the borders of the Empire such as \textit{Britannia} and \textit{Germania}. The high percentage of Dressel 20 amphorae found in sites nearby military garrisons suggests the existence of diverse supply dynamics as there were trade processes flowing from military sites to civil settlements\citep{remesal_annona_1986, carreras_britannia_1998}
\xavi{lo que quieres decir es que tienes muchas ánforas en sitios civiles cerca de militares? no queda claro}.
\maria{sí, pero creo que yo no he escrito está frase xD}

An explanation for the interest to export olive oil for the military garrison in the Roman provinces is still unclear. The origin of this exportation may be related to two aspects: a) cultural consumption whose olive oil is used by cultural reasons such as group identification or by habit and b) economical and political reasons to create a redistributive system mechanism to supply military camps.

%may be related to two aspects according with some hypotheses: a) cultural consumption whose product is consumed by cultural reasons such as to identity or habit and b) economical reasons where olive oil distribution is based on the transport costs, unlike other products \citep[69-70]{carreras_britannia_1998}\xavi{no pillo esta frase}..

@@

The evidence for olive oil consumption in \textit{Britannia} is scarce before the Roman conquest \citep{funari_corpus_1996,carreras_abastecimiento_2003}. Local population did not consume this product and it was not produced on the region as the landscape and climate of the British isles was not suitable for olive oil trees\citep[161]{monfort_britanniaen_1998}. Based on the archaeological records, this absence of olive oil changed with the arrival of a large amount of Dressel 20 amphorae to \textit{Britannia} from the first century A.D \citep{peacock_amphorae_1991,
carreras_britannia_1998}. At this moment, we detect an increase of the olive oil exportation concurring with the displacement of legions during the military campaigns \citep[161]{monfort_britanniaen_1998}
%\xavi{en qué siglo?}

The increase of the exportation of Dressel 20 amphorae created an important commercial network for exchanges. The demand of olive oil after the Roman conquest required of a new trade route due to the lack of local olive oil production; \textit{Baetica} was chosen as its main supply.

%At this moment, we detect an increase of the olive oil exportation concurring with the displacement of legions during the military campaigns \citep[161]{monfort_britanniaen_1998}\xavi{repetitivo? fusionar párrafos}.

This fact will have a particular intensity within sites close to the Hadrian Wall's garrisons. Olive oil production in \textit{Baetica} would cross the Atlantic until they reached the province and redistribute throughout the area from a series of strategic locations~\citep{carreras_atlantic_2012}.

%The increase of the exportation of Dressel 20 amphorae created an important commercial network for exchanges. Thus, the network was mainly focused on the support of soldiers during military campaigns\xavi{estas 2 frases repiten cosas ya explicadas}.

@@

According to previous studies those workshops became the most important centres of amphora production in the region \citep{rodriguez_economioleicola_1977,
arva_1997}.
%\xavi{cita de estos estudios}

It is worth mentioning that a majority of these stamps were collected during field surveys with no excavation, so this difference in sample size could be biased due to differences in intensity across the different sites~\citep{arva_1997}.

@@

The analysis of stamps in consumption areas also used the CEIPAC database and filtered in the same way than the previous dataset. However, in this case, we selected the centres with more or equal than five stamps in order to maximize the analysis with more stamps in the dataset.

%\xavi{pero esto no se hizo en el de Baetica no? Por qué?}.
